---
title: "RNAseq analysis including lymph node samples. CBR"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated analysis using clinical data from the Nature Cancer paper.

Keep lymph node samples for paired analysis between time points

2024-11-08: Added analysis for CBR

# R libraries and versions

```{r}
rm(list = ls())
library('readxl')
library('DESeq2')
library('ComplexHeatmap')
library(ggplot2)
library('EnhancedVolcano')
library('sva')
library('DT')
library('fgsea')
library('GSVA')
library(ggpubr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(msigdbr)
library(plyr)
library(apeglm)
library(xlsx)

set.seed(1234)
sessionInfo()

rootDir = "C:/Users/ldanilo1/OneDrive - Johns Hopkins/J15221_Evanthia_breast_data/"
#rootDir = "./"

source("params_and_functions.r")
# overwrite time colors
timeCol = setNames(c("black", 'grey45', 'grey75'), c("Baseline","C1D1","Week_8"))


```

# Load data
Load deidentified expression counts and sample annotation from the Nature Cancer paper

```{r echo=FALSE, message = FALSE, warning = FALSE}
exprsDat <- as.data.frame(read.table(paste0(rootDir,'Analysis/Data/RNAseq/j15221_evanthia_breast_rnaseq_deidentified.tsv'), row.names = 1, sep = '\t', header = T, check.names = F))

dim(exprsDat)  #[1] 22955    45

# annotation for samples that should be used for analysis
sampAnnot =read.csv(file=paste0(rootDir,"Analysis/06132023_updated_sampleAnnot_RNAseq.csv"), row.names = 1)

#pairs = printPairs(sampAnnot, patID = "Patient_ID", timepoint = "Timepoint")

#=======================
# load publication IDs
pubID = read.csv(paste0(rootDir,"Doc/publicationID.csv"), row.names = 1)
sampAnnot$pubID = pubID[sampAnnot$Patient_ID,1]
# publication timepoint labels
sampAnnot$pub_timepoint = factor(sampAnnot$Timepoint, levels = c("Baseline","C1D1","Week_8"), labels = c("Baseline","Post Run-in","Week 8"))
# save Personalis sample IDs
sampAnnot$PersonalisID = rownames(sampAnnot)
# change rownames to match expression data
rownames(sampAnnot) = paste(sampAnnot$pubID, sampAnnot$pub_timepoint, sep = "_")

#===================================
# load  patient table from the paper to update Best_resp
patTab = read.csv(paste0(rootDir,"for_paper/J15221_trial_paper/input_tables/table_24patient_info.csv"))
rownames(patTab) = patTab$ID
sampAnnot$Best_Resp = patTab[sampAnnot$pubID,"Best_Resp"]

# fix RECIST
sampAnnot$RECIST = factor(sampAnnot$Best_Resp, levels =c("PD","SD","CR","PR"), labels = c("NonResponder","NonResponder","Responder","Responder"))
# add CBR
sampAnnot$CBR = factor(sampAnnot$Best_Resp, levels = c("PD","SD","CR","PR"), labels = c("NonResponder","Responder","Responder","Responder"))

# output table
datatable(sampAnnot)

# save deidentified expression and the corresponding annotation object
# for further ananlysis
save(sampAnnot, exprsDat , file = paste0(rootDir,'Analysis/Data/RNAseq/j15221_rnaseq_annot_deidentified.rda'))

##Get VST for the whole dataset (including lymph nodes) for plotting
dds_plotting <- DESeqDataSetFromMatrix(countData = round(exprsDat[,rownames(sampAnnot)]), 
                              colData = sampAnnot, 
                              design=~Subtype)
vstDat_plotting <- assay(vst(dds_plotting))
vstDat_plotting <- vstDat_plotting[apply(vstDat_plotting,1,sd)>0,]

```

## Final sample numbers

```{r echo=FALSE, message = FALSE, warning = FALSE}
print("Sample Numbers")
sampAnnot %>% select(RECIST, Timepoint) %>% table()
# matrix for paired samples 
mat = printPairs(sampAnnot, timepoint = "Timepoint", patID = "Patient_ID")


table(sampAnnot$Time,sampAnnot$Subtype)

table(sampAnnot$Time,sampAnnot$CBR)


```

## Prep for gene set analysis

Loading HALLMARK gene sets, Evanthia's cell type markers, and lists of genes with strong and weak neoantigen binders for both class I and II

```{r echo=FALSE, message = FALSE, warning = FALSE}
#Gather Genes from msigdbr
HALLMARK_df = msigdbr(species = "Homo sapiens", category = "H")
HALLMARK_list = HALLMARK_df %>% split(x = .$gene_symbol, f = .$gs_name)
# load Evanthia's gene sets
# read in a table with mouse markers
markers = read.xlsx2(paste0(rootDir,"Doc/2023-05-16_mouse_paper/Human Cell Markers for scRNA-seq (1).xlsx"), 1, startRow = 2, endRow = 63)
# create list of gene sets
evanthia_list = markers %>% split(x = .$Representative.Gene.Markers, f = .$Cell.Type.)
Evanthia_CellTypes = lapply(evanthia_list,fixGeneNames)

# print Evanthia/Edgar's set of markers

print("top five Evanthia's gene sets:")
print(Evanthia_CellTypes[1:5])

# neoantigens
# lists of genes that predicted to have strong or weak binders
neoantigen_list = list()
neoantigen_list$class1_strong = unlist(read.table(paste0(rootDir,"Analysis/Data/Neoantigen/neoantigenGenes_classI_strong.txt")))
neoantigen_list$class1_weak = unlist(read.table(paste0(rootDir,"Analysis/Data/Neoantigen/neoantigenGenes_classI_weak.txt")))
neoantigen_list$class2_strong = unlist(read.table(paste0(rootDir,"Analysis/Data/Neoantigen/neoantigenGenes_classII_strong.txt")))
neoantigen_list$class2_weak = unlist(read.table(paste0(rootDir,"Analysis/Data/Neoantigen/neoantigenGenes_classII_weak.txt")))

print("Neoantigens:")
print(sapply(neoantigen_list, length))
```

# CBR DE

For this DE, we will be doing a DE analysis to compare CBR responder v nonresponder.

## Clinical Responder vs Clinical Nonresponder (Baseline)

```{r echo=FALSE, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="Baseline",]
table(annot$CBR)

res = getDEres(round(exprsDat[,rownames(annot)]), 
               annot, "~Tissue_Broad+CBR", 
               contrast = "CBR_Responder_vs_NonResponder", lfcShrink = T)

# output all DE results  
datatable(data.frame(res))
# plots volcano plot and saves top DE genes
topDE = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_DEGenes.csv"),
                       main_title = "Baseline. NonResponder vs Responder (CBR)")

# print some statistics
cat("The number of significantly DE genes:\n",
length(topDE))
cat("The number of down- and up-regulated genes:\n",
    table(res[topDE,"log2FoldChange"] >0))

```

### Heatmap of Responder vs Nonresponder
```{r echo=FALSE, message = FALSE, warning = FALSE, fig.height=7}
##use all samples including lymph nodes
plotHA_cbr(sampAnnot[sampAnnot$Timepoint=="Baseline",], vstDat_plotting, topDE)

```

### Hallmark Pathway analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}
  stats <- res$log2FoldChange
  names(stats) <- row.names(res)

##Pathway Analysis
baseline_CBR_NonRespondervResponder_gsResults <-
  list(HALLMARK=fgsea(pathways=HALLMARK_list, stats=stats),
       Evanthia_CellTypes=fgsea(pathways=Evanthia_CellTypes, stats=stats),
       neoantigens = fgsea(pathways=neoantigen_list, stats=stats))

datatable(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK)
fwrite(as.data.frame(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_Hallmark.csv"))

# make barplots
fgseaBarplot(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK, 
             main_title = "Hallmark Pathways. Baseline. NonResponder vs Responder (CBR)")

```

### Evanthia_Celltypes gene set analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}
datatable(baseline_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes)
fwrite(as.data.frame(baseline_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_Evanthia_CellTypes.csv"))

# make barplots
fgseaBarplot(baseline_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes, 
             main_title = "Evanthia_CellTypes. Baseline. NonResponder vs Responder (CBR)")

```

### Neoantigen lists

```{r echo=FALSE, message = FALSE, warning = FALSE}
# intersect with topDE genes
print("Intersect with top DE genes")
sapply(neoantigen_list, function(x) length(intersect(x,topDE)))

datatable(baseline_CBR_NonRespondervResponder_gsResults$neoantigens)
fwrite(as.data.frame(baseline_CBR_NonRespondervResponder_gsResults$neoantigens), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_neoantigens.csv"))

```

## Responder vs Nonresponder (C1D1)

```{r echo=FALSE, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
table(annot$CBR)

res = getDEres(round(exprsDat[,rownames(annot)]), 
               annot, "~Tissue_Broad+CBR", 
               contrast = "CBR_Responder_vs_NonResponder", lfcShrink = T)

# output all DE results  
datatable(data.frame(res))
# plots volcano plot and saves top DE genes
topDE = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_DEGenes.csv"),
                       main_title = "C1D1. NonResponder vs Responder (CBR)")

# print some statistics
cat("The number of significantly DE genes:\n",
length(topDE))
cat("The number of down- and up-regulated genes:\n",
    table(res[topDE,"log2FoldChange"] >0))
```

### Heatmap of Responder vs Nonresponder

```{r echo=FALSE, message = FALSE, warning = FALSE, fig.height=7}
##use all samples including lymph nodes
anno <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
plotHA_cbr(anno, vstDat_plotting, topDE)

```

### Hallmark Pathway analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}
  stats <- res$log2FoldChange
  names(stats) <- row.names(res)

##Pathway Analysis
C1D1_CBR_NonRespondervResponder_gsResults <-
  list(HALLMARK=fgsea(pathways=HALLMARK_list,stats=stats),
       Evanthia_CellTypes=fgsea(pathways=Evanthia_CellTypes,stats=stats),
       neoantigens = fgsea(pathways=neoantigen_list, stats=stats))

  datatable(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK)
fwrite(as.data.frame(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_Hallmark.csv"))

# make barplots
fgseaBarplot(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK, 
             main_title = "Hallmark pathways. C1D1. NonResponder vs Responder (CBR)")

```

### Evanthia_Celltypes gene set analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}

datatable(C1D1_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes)
fwrite(as.data.frame(C1D1_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_Evanthia_CellTypes.csv"))

# make barplots
fgseaBarplot(C1D1_CBR_NonRespondervResponder_gsResults$Evanthia_CellTypes, 
             main_title = "Evanthia_CellTypes. C1D1. NonResponder vs Responder (CBR)")

```

### Neoantigen lists

```{r echo=FALSE, message = FALSE, warning = FALSE}
# intersect with topDE genes
print("Intersect with top DE genes")
sapply(neoantigen_list, function(x) length(intersect(x,topDE)))

datatable(C1D1_CBR_NonRespondervResponder_gsResults$neoantigens)
fwrite(as.data.frame(C1D1_CBR_NonRespondervResponder_gsResults$neoantigens), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_neoantigens.csv"))

```

## Responder vs Nonresponder (Week 8)

Not enough samples to do DE analysis

