---
title: "RNAseq analysis including lymph node samples. CBR"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
library('readxl')
library('DESeq2')
library('ComplexHeatmap')
library(ggplot2)
library('EnhancedVolcano')
library('sva')
library('DT')
library('fgsea')
library('GSVA')
library(ggpubr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(msigdbr)
library(plyr)
library(apeglm)
library(xlsx)

set.seed(1234)
sessionInfo()

rootDir = "C:/Users/Luda/OneDrive - Johns Hopkins/J15221_Evanthia_breast_data/"
#rootDir = "./"

source("params_and_functions.r")
# overwrite time colors
timeCol = setNames(c("black", 'grey45', 'grey75'), c("Baseline","C1D1","Week_8"))


```

Updated analysis using clinical data from the Nature Cancer paper.

Keep lymph node samples for paired analysis between time points

# Load data
Load deidentified expression counts and sample annotation from the Nature Cancer paper

```{r echo=FALSE, message = FALSE, warning = FALSE}
# load deidentified expression and the corresponding annotation object
load(file = paste0(rootDir,'Analysis/Data/RNAseq/j15221_rnaseq_annot_deidentified.rda'))

dim(exprsDat)  #[1] 22955    37

# output table
datatable(sampAnnot)

dds_plotting <- DESeqDataSetFromMatrix(countData = round(exprsDat[,rownames(sampAnnot)]), 
                              colData = sampAnnot, 
                              design=~Subtype)
vstDat_plotting <- assay(vst(dds_plotting))
vstDat_plotting <- vstDat_plotting[apply(vstDat_plotting,1,sd)>0,]

```

## Final sample numbers

```{r echo=FALSE, message = FALSE, warning = FALSE}
print("Sample Numbers")
sampAnnot %>% select(RECIST, Timepoint) %>% table()
# matrix for paired samples 
mat = printPairs(sampAnnot, timepoint = "Timepoint", patID = "Patient_ID")


table(sampAnnot$Time,sampAnnot$Subtype)

table(sampAnnot$Time,sampAnnot$CBR)


```

## Prep for gene set analysis

Loading HALLMARK gene sets, Evanthia's cell type markers, and lists of genes with strong and weak neoantigen binders for both class I and II

```{r echo=FALSE, message = FALSE, warning = FALSE}
#Gather Genes from msigdbr
HALLMARK_df = msigdbr(species = "Homo sapiens", category = "H")
HALLMARK_list = HALLMARK_df %>% split(x = .$gene_symbol, f = .$gs_name)
```

# CBR DE

For this DE, we will be doing a DE analysis to compare CBR responder v nonresponder.

## Clinical Responder vs Clinical Nonresponder (Baseline)

```{r echo=FALSE, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="Baseline",]
table(annot$CBR)

res = getDEres(round(exprsDat[,rownames(annot)]), 
               annot, "~Tissue_Broad+CBR", 
               contrast = "CBR_Responder_vs_NonResponder", lfcShrink = T)

# output all DE results  
datatable(data.frame(res))
# plots volcano plot and saves top DE genes
topDE = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_DEGenes.csv"),
                       main_title = "Baseline. NonResponder vs Responder (CBR)")

```

### Heatmap of Responder vs Nonresponder
```{r echo=FALSE, message = FALSE, warning = FALSE, fig.height=7}
##use all samples including lymph nodes
plotHA_cbr(sampAnnot[sampAnnot$Timepoint=="Baseline",], vstDat_plotting, topDE)

```

### Hallmark Pathway analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}
  stats <- res$log2FoldChange
  names(stats) <- row.names(res)

##Pathway Analysis
baseline_CBR_NonRespondervResponder_gsResults <-
  list(HALLMARK=fgsea(pathways=HALLMARK_list, stats=stats))

datatable(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK)
fwrite(as.data.frame(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/Baseline_CBR_RespondervNonResponder_Hallmark.csv"))

# make barplots
fgseaBarplot(baseline_CBR_NonRespondervResponder_gsResults$HALLMARK, 
             main_title = "Hallmark Pathways. Baseline. NonResponder vs Responder (CBR)")

```


## Responder vs Nonresponder (C1D1)

```{r echo=FALSE, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
table(annot$CBR)

res = getDEres(round(exprsDat[,rownames(annot)]), 
               annot, "~Tissue_Broad+CBR", 
               contrast = "CBR_Responder_vs_NonResponder", lfcShrink = T)

# output all DE results  
datatable(data.frame(res))
# plots volcano plot and saves top DE genes
topDE = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_DEGenes.csv"),
                       main_title = "C1D1. NonResponder vs Responder (CBR)")

```

### Heatmap of Responder vs Nonresponder

```{r echo=FALSE, message = FALSE, warning = FALSE, fig.height=7}
##use all samples including lymph nodes
anno <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
plotHA_cbr(anno, vstDat_plotting, topDE)

```

### Hallmark Pathway analysis
```{r echo=FALSE, message = FALSE, warning = FALSE}
  stats <- res$log2FoldChange
  names(stats) <- row.names(res)

##Pathway Analysis
C1D1_CBR_NonRespondervResponder_gsResults <-
  list(HALLMARK=fgsea(pathways=HALLMARK_list,stats=stats))

  datatable(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK)
fwrite(as.data.frame(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK), file=paste0(rootDir,"Analysis/Data/RNAseq/2024-01-22_LN/C1D1_CBR_RespondervNonResponder_Hallmark.csv"))

# make barplots
fgseaBarplot(C1D1_CBR_NonRespondervResponder_gsResults$HALLMARK, 
             main_title = "Hallmark pathways. C1D1. NonResponder vs Responder (CBR)")

```

## Responder vs Nonresponder (Week 8)

Not enough samples to do DE analysis

