---
title: "Molecular subtypes, TCR, and neoantigens"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(ggplot2)
library('DT')
library(ggpubr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(plyr)
library(kableExtra)
library(RColorBrewer)
library(ComplexHeatmap)

set.seed(1234)
sessionInfo()

inputFolder = "input_data/"
outputFolder = "output_data/"
# check if it exists, if not, create
if (!dir.exists(outputFolder)) {
  dir.create(outputFolder)
}
# load all necessary functions and settings

source("params_and_functions.r")
```


# Load data
Load patient information from the Nature Cancer paper repo (https://github.com/FertigLab/J15221_trial_paper).

```{r echo=FALSE, message = FALSE, warning = FALSE}
patInfoLink = "https://raw.githubusercontent.com/FertigLab/J15221_trial_paper/refs/heads/main/input_tables/table_24patient_info.csv"

patInfo = read.csv(patInfoLink, row.names = 1, stringsAsFactors = F)

# read sample into and merge with patient info
sampAnnot = read.csv(paste0(inputFolder,"diversity_neoantigen_table.csv"), row.names = 1)

sampAnnot = cbind(sampAnnot, patInfo[sampAnnot$patientID, c("Best_Resp", "biopsy_site")])

# add timepoint
sampAnnot = sampAnnot %>% mutate(Timepoint = Time)

```
# Molecular subtypes

PAM50 predictions were done using the pam50 R package.
https://github.com/hoadley-lab/pam50
Parker et al. 2009, https://pubmed.ncbi.nlm.nih.gov/19204204/

The results of predictions are in the `input_data/PAM50` folder 

## TNBCtype assingments

To classify samples according to TNBCtype subtypes, we uploaded the gene expression data to the TNBCtype webtool (https://cbc.app.vumc.org/tnbc/). Out of 37 samples, the tool was able to classify 13 samples from six TNBC patients and one week 8 sample from HR+ patient R-9. 
 
The results of predictions are in the `input_data/TNBCtype` folder


```{r echo=FALSE, fig.width = 10, fig.height = 10, message = FALSE, warning = FALSE}
# add PAM50 assignments to the sample annotation
# read in pam50 assignments
pam50_assign = read.table(paste0(inputFolder,'PAM50/pam50_assignments.tsv'), sep = '\t', header = T, row.names = 1)
# fix the row names
rownames(pam50_assign) = gsub("Post Run-in","C1D1",rownames(pam50_assign))
rownames(pam50_assign) = gsub("Week 8","Week_8",rownames(pam50_assign))

sampAnnot_pam50 = merge(sampAnnot, pam50_assign, by = "row.names", all.x = T)
rownames(sampAnnot_pam50) = sampAnnot_pam50$Row.names

#=========================
# add TNBCtype assignments to the sample annotation
# read in TNBCtype assignments
tnbc_assign = read.csv(paste0(inputFolder,'TNBCtype/5ccca5bc-a76e-4320-9c01-49817ade4087_result.csv'), header = T, row.names = 1)
# coefficients
tnbc_coef = read.csv(paste0(inputFolder,'TNBCtype/5ccca5bc-a76e-4320-9c01-49817ade4087_coef.csv'), header = T, row.names = 1)

# add this assignment to the patient annotation
# but first we need to fix samples names in tnbc_assign
rownames(tnbc_assign) = fix_sample_names(rownames(tnbc_assign))
rownames(tnbc_coef) = fix_sample_names(rownames(tnbc_coef))

datatable(tnbc_assign, options = list(pageLength = 13))
datatable(round(tnbc_coef,3), options = list(pageLength = 13))

# add to sample and per patient annotation
sampAnnot_pam50$TNBCtype = "NA"
s = intersect(rownames(tnbc_assign), rownames(sampAnnot_pam50))
sampAnnot_pam50[s,"TNBCtype"] = tnbc_assign[s,"subtype"]

perPat = sampAnnot_pam50[!duplicated(sampAnnot_pam50$patientID),]
s = intersect(rownames(tnbc_assign), rownames(perPat))
perPat$TNBCtype = "NA"
perPat[s,"TNBCtype"] = tnbc_assign[s,"subtype"]

```

## Heatmap

The heatmap was generated using the PAM50 genes were selected from the `pam50` package. The heatmap was clustered by samples and genes. The color scheme was set to blue for low expression and red for high expression. The annotation was added to the left side of the heatmap, showing the PAM50 subtype, RECIST response, best response, biopsy site, patient ID, and time point.


```{r echo=FALSE, fig.width = 10, fig.height = 10, message = FALSE, warning = FALSE}

# get pam50 genes
genes = pam50::pam50_annot$gene_name

# read in data for heatmap
dat = read.csv(paste0(inputFolder,"PAM50/PAM50_genes_heatmap.csv"), row.names = 1, check.names = F)

ord = rownames(sampAnnot_pam50)[order(sampAnnot_pam50$Subtype,sampAnnot_pam50$RECIST,sampAnnot_pam50$Best_Resp, sampAnnot_pam50$biopsy_site, sampAnnot_pam50$patientID, sampAnnot_pam50$Timepoint)]
sampAnnot_pam50 = sampAnnot_pam50[ord,]

ht = plotHA_pam50_tnbc(sampAnnot_pam50, dat, genes, title = "All samples",
       column_labels = sampAnnot_pam50$patientID, show_row_names = T,
       cluster_columns = F)


```

## PAM50 correlation with response

Assign per patient PAM50 subtype to correlate with response. If a patient doesn't have a baseline samples, use subsequent time points.


```{r echo=FALSE, fig.width = 8, fig.height = 12, message = FALSE, warning = FALSE}

datatable(perPat[,c("patientID","RECIST","Best_Resp","biopsy_site","Timepoint","assignment","TNBCtype")],
          options = list(pageLength = 19))

tab = table(perPat$assignment, perPat$RECIST)
print(tab)
chisq.test(tab)

tab = table(perPat$assignment, perPat$Subtype)
print(tab)
chisq.test(tab)


```

# TCR diversity
T cell receptor repertoires derived from RNA sequencing data were estimated using MiXCR and provided by Personalis Inc. (Menlo Park, CA, USA). We loaded the provided data into R programming environment using the immunarch package and calculated the diversity using the repDiversity function with the chaos1 method. The diversity was visualized as boxplots with pairwise two-sided Wilcoxon rank-sum test p-values. 

```{r echo=FALSE, fig.width=10, eval=TRUE, message = FALSE, warning = FALSE}
df_all = CreateAllFacet(sampAnnot,"RECIST")
p = ggpaired(df_all , x="Time",y="chaos1", id = "patientID",
             line.size = 0.8, ylab = "Diversity",
             line.color = "RECIST",
             palette =  recistCol, xlab="Time",
             ggtheme = theme_minimal()) +
        facet_wrap("facet") +
        stat_compare_means(
          comparisons = list(c(1,2), c(2,3),c(1,3)), size = 5)+
        scale_x_discrete(labels=timeLab)+
  theme(text = element_text(size = 20),
        axis.text.x = element_text(size = 17), legend.position="none")
  print(p)

 

```

# Strong binding neoantigens
Neoantigen binding affinity was estimated using NetMHCIIPan 4.0 by Personalis Inc. (Menlo Park, CA, USA). To calculate the number of strong binders per sample, we took only peptides of expressed genes, and then for every peptide, we took the minimal predicted HLA rank. After that, we calculated the number of peptides that have rank less than 2 as recommended by Reynisson et al (https://academic.oup.com/nar/article/48/W1/W449/5837056) to be a strong binder. 

```{r echo=FALSE, fig.width=10, fig.height=6 , eval=TRUE, message = FALSE, warning = FALSE}

 p =  ggpaired(CreateAllFacet(sampAnnot, "RECIST") , x="Time",y="nNeoantigens_strong", id = "patientID", line.size = 0.8, 
               line.color = "RECIST",palette =  recistCol,
               ggtheme = theme_minimal(),
               title = "", xlab="Time", ylab = "The number of strong binders") +
      facet_wrap("facet") +
      scale_x_discrete(labels=timeLab)+
      scale_y_continuous(trans = "log10")+
      stat_compare_means(comparisons =list(c(1,2), c(2,3),c(1,3)), size = 5)+
  theme(text = element_text(size = 20),
        axis.text.x = element_text(size = 17), legend.position="none") 

 print(p)
```

# R libraries and versions

```{r echo=FALSE, message = FALSE, warning = FALSE}

sessionInfo()

```

