---
title: "Visualization of RNAseq data analysis"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library('ComplexHeatmap')
library(ggplot2)
library('EnhancedVolcano')
library('DT')
library(ggpubr)
library(ggfortify)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(plyr)
library(kableExtra)
library(RColorBrewer)

set.seed(1234)

inputFolder = "input_data/"
outputFolder = "output_data/"
# check if it exists, if not, create
if (!dir.exists(outputFolder)) {
  dir.create(outputFolder)
}
# load all necessary functions and settings
source("params_and_functions.r")

```

Visualization of saved results from the RNAseq DE analysis. 

# R libraries and versions

```{r echo=FALSE, message = FALSE, warning = FALSE}

sessionInfo()

```

# Load data
Load patient annotation from the repository of the previously published Nature Cancer paper: https://github.com/FertigLab/J15221_trial_paper.

```{r echo=FALSE, message = FALSE, warning = FALSE}
patInfoLink = "https://raw.githubusercontent.com/FertigLab/J15221_trial_paper/refs/heads/main/input_tables/table_24patient_info.csv"

patInfo = read.csv(patInfoLink, row.names = 1, stringsAsFactors = F)

# read sample into and merge with patient info
sampAnnot = read.csv(paste0(inputFolder,"diversity_neoantigen_table.csv"), row.names = 1)

sampAnnot = cbind(sampAnnot, patInfo[sampAnnot$patientID, c("Best_Resp", "biopsy_site")])

# add timepoint
sampAnnot = sampAnnot %>% mutate(Timepoint = Time)

# add pubID
sampAnnot$pubID = sapply(strsplit(rownames(sampAnnot), "_"),
                            getElement, 1)

```

## Sample numbers

```{r echo=FALSE, message = FALSE, warning = FALSE}
print("Sample Numbers")

tab = sampAnnot %>% select(RECIST, Timepoint) %>% table()
tab = rbind(tab, colSums(tab))
kable(tab,"simple")
# matrix for paired samples 
mat = printPairs(sampAnnot, timepoint = "Timepoint", patID = "patientID")

#table(sampAnnot$biopsy_site)

#table(sampAnnot$Time)

#table(sampAnnot$biopsy_site,sampAnnot$Subtype)

# tab = table(sampAnnot$biopsy_site,sampAnnot$Time)
# tab = cbind(tab, rowSums(tab))
# kable(tab,"simple")


# tab = table(sampAnnot$biopsy_site,sampAnnot$RECIST)
# tab = cbind(tab, rowSums(tab))
# kable(tab,"simple")

table(sampAnnot$Subtype,sampAnnot$RECIST)

table(sampAnnot$Time,sampAnnot$RECIST)

table(sampAnnot$Time,sampAnnot$Subtype)

```

# Treatment DE

Paired DE analysis between treatments/timepoints.

## Baseline vs C1D1

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
# find patients with paired samples
mat = printPairs(sampAnnot, timepoint = "Timepoint", patID = "patientID")
pairedPat <- mat %>% filter(Baseline == 1 & C1D1 == 1) %>% rownames()
# select paired samples and exclude a time point from comparison
annot <- sampAnnot %>% filter(patientID %in% pairedPat & Timepoint != "Week_8")

##read in results
res = read.csv(paste0(inputFolder,"DE_treatment/Baseline_vs_C1D1_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_b_vs_c1 = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "Baseline vs C1D1", n = 13)


```

### Heatmap
```{r echo=FALSE, fig.width=7, fig.height = 6, message = FALSE, warning = FALSE}

# read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_treatment/Baseline_vs_C1D1_DEGenes_heatmap.csv"), row.names = 1, check.names = F)


ht = plotHA(annot, dat, rownames(dat), scale = F,
       title = "Baseline vs C1D1",  
       column_labels = annot$pubID, show_row_names = F,
       heatmap_legend_param = list(direction = "horizontal"))


```


## C1D1 vs Week_8

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
# find patients with paired samples
pairedPat <- mat %>% filter(Week_8 == 1 & C1D1 == 1) %>% rownames()
# select paired samples and exclude a time point from comparison
annot <- sampAnnot %>% filter(patientID %in% pairedPat & Timepoint != "Baseline")

#datatable(annot)

##DE by Time
res = read.csv(paste0(inputFolder,"DE_treatment/Week_8_vs_C1D1_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_c_vs_w8 = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "C1D1 vs Week 8", n = 15)

```

### Heatmap

```{r echo=FALSE, fig.width=7, fig.height = 6, message = FALSE, warning = FALSE}
# read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_treatment/C1D1_vs_Week8_DEGenes_heatmap.csv"), row.names = 1, check.names = F)

plotHA(annot, dat, rownames(dat), title = "C1D1 vs Week 8",
       scale = F, column_labels = annot$pubID, show_row_names = F,
       heatmap_legend_param = list(direction = "horizontal"))

```


## Baseline vs Week_8

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
# find patients with paired samples
pairedPat <- mat %>% filter(Week_8 == 1 & Baseline == 1) %>% rownames()
# select paired samples and exclude a time point from comparison
annot <- sampAnnot %>% filter(patientID %in% pairedPat & Timepoint != "C1D1")
#datatable(annot)

res = read.csv( paste0(inputFolder,"DE_treatment/Week_8_vs_Baseline_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_b_vs_w8 = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "Baseline vs Week 8")

```

### Heatmap

```{r echo=FALSE, fig.width=7,fig.height = 6, message = FALSE, warning = FALSE}
# read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_treatment/Baseline_vs_Week8_DEGenes_heatmap.csv"), row.names = 1, check.names = F)


plotHA(annot, dat, rownames(dat), 
       title = "Baseline vs Week 8",scale = F,
       column_labels = annot$pubID, show_row_names = F,
       heatmap_legend_param = list(direction = "horizontal"))

```

## Hallmark Pathway analysis
### dot plot

```{r echo=FALSE, message = FALSE, warning = FALSE,fig.height=7, fig.width=5}
# load pathways from paired analysis
c1d1_week8_h = read.csv(paste0(inputFolder,"DE_treatment/Week_8_vs_C1D1_Hallmark.csv"))
c1d1_week8_h_sig = c1d1_week8_h %>% filter(padj < 0.05)

b_week8_h = read.csv(paste0(inputFolder,"DE_treatment/Week_8_vs_Baseline_Hallmark.csv"))
b_week8_h_sig = b_week8_h %>% filter(padj < 0.05)

# merge pathwyay analysis results for plotting
pathRes = rbind(cbind(comparison = rep("Baseline_vs_Week8",nrow(b_week8_h)),b_week8_h),
  cbind(comparison = rep("C1D1_vs_Week8",nrow(c1d1_week8_h)),c1d1_week8_h))

 # make NES for not significant pathways as NA
pathRes_ns = pathRes %>% mutate(NES, NES = ifelse(padj >0.05, NA, NES))
# remove "HALLMARK_" from all pathways
pathRes_ns$pathway = gsub("HALLMARK_","",pathRes_ns$pathway)

# plot: dot plot
p = ggplot(data = pathRes_ns %>% filter(!is.na(NES)), aes(x = comparison, y = pathway, color = NES)) + 
    geom_point(size = 4) +
  #  scale_color_gradient(low = "red", high = "blue") +
    scale_colour_gradient2(low = "blue", high = "red") +
  scale_x_discrete(labels = c("Baseline vs Week 8","C1D1 vs Week 8")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle("Hallmark pathways") 


  print(p)

```


# RECIST DE

Results of DE analysis between responders vs non-responders at each timepoint.

## Baseline

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="Baseline",]
table(annot$RECIST)

res = read.csv(paste0(inputFolder,"DE_RECIST/Baseline_RespondervNonResponder_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_b = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "Baseline. Non-responder vs Responder")

# create a table with top 10 or less up- and down-regulated genes for every comparison
top10Genes = matrix(NA, nrow = 10, ncol = 6)
colnames(top10Genes) = c("Baseline_up","Baseline_down","C1D1_up","C1D1_down",
                         "Week8_up","Week8_down")
# add top gene from this comparison
sig = res[topDE_b,"log2FoldChange"] >0
top10Genes[,"Baseline_up"] = topDE_b[sig][1:10]
top10Genes[,"Baseline_down"] = topDE_b[!sig][1:10]
                                          
```

### top DE genes

```{r echo=FALSE, fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
##read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_RECIST/Baseline_R_vs_NR_DEGenes_heatmap.csv"), row.names = 1, check.names = F)
annot = sampAnnot[sampAnnot$Timepoint=="Baseline",]
##use all samples including lymph nodes
plotHA(annot, dat, rownames(dat), title = "Baseline",scale = F,
       column_labels = annot$pubID, show_row_names = F,
      heatmap_legend_param = list(direction = "horizontal"))

```

### top Hallmark pathways

Pathway scores for significantly enriched pathways derived for each sample using GSVA

```{r echo=FALSE, fig.width=8, fig.heigh = 6, message = FALSE, warning = FALSE, eval = TRUE}
# read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_RECIST/Baseline_R_vs_NR_pathways_heatmap.csv"), row.names = 1, check.names = F)

annot =sampAnnot %>% filter(Timepoint == "Baseline") 
# remove "HALLMARK_" from all pathway names to be used as row names
p = gsub("HALLMARK_","",rownames(dat))
plotHA(annot, dat, scale = F,
           title = "Hallmark pathways. Baseline. GSVA scores", col = heatmapCol,column_labels = annot$pubID,
           row_labels = p)


```

## C1D1

```{r echo=FALSE, fig.width = 6, fig.height=6,message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
table(annot$RECIST)

res = read.csv(paste0(inputFolder,"DE_RECIST/C1D1_RespondervNonResponder_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_c = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "C1D1. Non-responder vs Responder")
# add top gene from this comparison
sig = res[topDE_c,"log2FoldChange"] >0
top10Genes[,"C1D1_up"] = topDE_c[sig][1:10]
top10Genes[,"C1D1_down"] = topDE_c[!sig][1:10]

```

### Heatmap 

```{r echo=FALSE,fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
##read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_RECIST/C1D1_R_vs_NR_DEGenes_heatmap.csv"), row.names = 1, check.names = F)

annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
plotHA(annot, dat, rownames(dat),title = "C1D1",scale = F,
       column_labels = annot$pubID, show_row_names = F,
       heatmap_legend_param = list(direction = "horizontal"))

```

## Week 8

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="Week_8",]
table(annot$RECIST)

res = read.csv(paste0(inputFolder,"DE_RECIST/Week8_RespondervNonResponder_DEGenes.csv"), row.names = 1)
# output all DE results  
#datatable(data.frame(res))
# plots volcano plot and saves top DE genes
topDE_w8 = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "Week 8. Non-responder vs Responder", n = 25)
# add top gene from this comparison
sig = res[topDE_w8,"log2FoldChange"] >0
top10Genes[,"Week8_up"] = topDE_w8[sig][1:10]
top10Genes[,"Week8_down"] = topDE_w8[!sig][1:10]


```

### Heatmap

```{r echo=FALSE, fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
##read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_RECIST/Week8_R_vs_NR_DEGenes_heatmap.csv"), row.names = 1, check.names = F)

annot <- sampAnnot[sampAnnot$Timepoint=="Week_8",]
plotHA(annot, dat, rownames(dat), title = "Week 8",scale = F,
       column_labels = annot$pubID, show_row_names = F,
       heatmap_legend_param = list(direction = "horizontal"))


```


## Top 10 gene table

```{r echo=FALSE, fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
datatable(top10Genes)
write.csv(top10Genes,paste0(outputFolder,"top10Genes.csv"), row.names = F)

```


# CBR DE

Results of DE analysis between CBR responders vs non-responders at each timepoint.

## CBR numbers

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
# add CBR
sampAnnot$CBR = factor(sampAnnot$Best_Resp, levels = c("PD","SD","CR","PR"), labels = c("Non-responder","Responder","Responder","Responder"))

table(sampAnnot$Timepoint,sampAnnot$CBR)

```

## Baseline

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="Baseline",]
table(annot$RECIST)

res = read.csv(paste0(inputFolder,"DE_CBR/Baseline_CBR_RespondervNonResponder_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_b = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "Baseline. CBR Non-responder vs Responder")

```
### Heatmap

```{r echo=FALSE, fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
##read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_CBR/Baseline_CBR_R_vs_NR_DEGenes_heatmap.csv"), row.names = 1, check.names = F)

annot <- sampAnnot[sampAnnot$Timepoint=="Baseline",]
plotHA(annot, dat, rownames(dat), title = "Baseline",scale = F,
       column_labels = annot$pubID, show_row_names = T)


```

## C1D1

```{r echo=FALSE, fig.width = 6, fig.height=6, message = FALSE, warning = FALSE}
##Subset samples to be included
annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
table(annot$RECIST)

res = read.csv(paste0(inputFolder,"DE_CBR/C1D1_CBR_RespondervNonResponder_DEGenes.csv"), row.names = 1)

# output all DE results  
datatable(data.frame(res))

# plots volcano plot and saves top DE genes
topDE_b = plotVolcano(res, fdrThr = 0.05, logFCthr = 0.5, 
                       fileName = NULL,
                       main_title = "C1D1. CBR Non-responder vs Responder")

```
### Heatmap

```{r echo=FALSE, fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
##read in data for heatmap
dat = read.csv(paste0(inputFolder,"DE_CBR/C1D1_CBR_R_vs_NR_DEGenes_heatmap.csv"), row.names = 1, check.names = F)

annot <- sampAnnot[sampAnnot$Timepoint=="C1D1",]
plotHA(annot, dat, rownames(dat), title = "C1D1",scale = F,
       column_labels = annot$pubID, show_row_names = T)


```
