---
title: "Visualization of RNAseq data analysis"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library('readxl')
library('DESeq2')
library('ComplexHeatmap')
library(ggplot2)
library('EnhancedVolcano')
library('DT')
library(ggpubr)
library(ggfortify)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(msigdbr)
library(plyr)
library(apeglm)
library(xlsx)
library(kableExtra)
library(pam50)
library(RColorBrewer)
library(GSVA)
library(readr)

set.seed(1234)

inputFolder = "input_data/"
outputFolder = "output_data/"
# check if it exists, if not, create
if (!dir.exists(outputFolder)) {
  dir.create(outputFolder)
}

source("params_and_functions.r")

# p-value threshold
pThresh = 0.1

```

Visualization for the paper using saved results from the RNAseq analysis. 

# R libraries and versions

```{r echo=FALSE, message = FALSE, warning = FALSE}

sessionInfo()

```

# Load data
Load sample annotation from the Nature Cancer paper repo (https://github.com/FertigLab/J15221_trial_paper).

```{r echo=FALSE, message = FALSE, warning = FALSE}
patInfoLink = "https://raw.githubusercontent.com/FertigLab/J15221_trial_paper/refs/heads/main/input_tables/table_24patient_info.csv"

patInfo = read.csv(patInfoLink, row.names = 1, stringsAsFactors = F)

# read sample into and merge with patient info
sampAnnot = read.csv(paste0(inputFolder,"diversity_neoantigen_table.csv"), row.names = 1)

sampAnnot = cbind(sampAnnot, patInfo[sampAnnot$patientID, c("Best_Resp", "biopsy_site")])

# add timepoint
sampAnnot$Timepoint = sapply(strsplit(rownames(sampAnnot), "_"),
                            getElement, 2)

```

# CIBERSORTx

The analysis was performed using the CIBERSORTx web portal (https://cibersortx.stanford.edu/) with the LM22 signature matrix. The results were downloaded and processed in R.


```{r echo=FALSE, fig.width = 8, message = FALSE, warning = FALSE}
#=============================== 
# read in CIBERSORT proportions
ciberRes = read.csv(file = paste0(rootDir,'Analysis/Data/cibersort_results/CIBERSORTx_Job25_Results.csv'), row.names = 1)

# subset to samples that we agreed to use for further analysis
ciberRes = ciberRes[sampAnnot$PersonalisID,1:22]

```
# TME gene sets
## Significant sets in responders vs non-responder at C1D1

```{r echo=FALSE, fig.width = 8, message = FALSE, warning = FALSE}
# read in scores
scores = read.xlsx(paste0(rootDir,"Doc/2024-11-04_RNAseq_paper/Supplementary_tables_15-19_TME.xlsx"), sheetIndex = 2, startRow = 2, colNames = T, check.names = F)
rownames(scores) = scores[,1]
scores = scores[,-1]

# read in p-values
res = read.xlsx(paste0(rootDir,"Doc/2024-11-04_RNAseq_paper/Supplementary_tables_15-19_TME.xlsx"), sheetIndex = 3, startRow = 2, colNames = T, check.names = F)
rownames(res) = res[,1]
res = res[,-1]
sig_t1 = rownames(res)[which(res[,"pValue_T1_RvsNR"] < pThresh)]


# order by patient and time
ord_all = rownames(sampAnnot)[order(sampAnnot$RECIST,sampAnnot$Best_Resp, sampAnnot$Subtype, sampAnnot$pubID,sampAnnot$Timepoint)]
# C1D1 only
ord = grep("Post Run-in",ord_all, value = T)
# heatmap
plotHA(sampAnnot[ord,],scores[sig_t1,ord], scale = F, title = "TME gene sets. C1D1", col = heatmapCol, column_order = ord, column_labels  = gsub("_Post Run-in","",ord))

```

## Significant sets in responders vs non-responder at Week 8

```{r echo=FALSE, fig.width=8, eval=TRUE}
# get significant by time point
sig_t2 = rownames(res)[which(res[,"pValue_T2_RvsNR"] < pThresh)]


# Week 8 only
ord = grep("Week 8",ord_all, value = T)
plotHA(sampAnnot[ord,],scores[sig_t2,ord], scale = F, title = "TME gene sets. Week 8", col = heatmapCol, column_order = ord,
          column_labels  = gsub("_Week 8","",ord) )


```

