---
title: "Visualization of TME analysis"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library('readxl')
library('ComplexHeatmap')
library(ggplot2)
library('DT')
library(ggpubr)
library(ggfortify)
library(tidyverse)
library(dplyr)
library(tidyr)
library(data.table)
library(plyr)
library(xlsx)
library(kableExtra)
library(pam50)
library(RColorBrewer)

set.seed(1234)

inputFolder = "input_data/"
outputFolder = "output_data/"
# check if it exists, if not, create
if (!dir.exists(outputFolder)) {
  dir.create(outputFolder)
}
# load all necessary functions and settings

source("params_and_functions.r")

# p-value threshold
pThresh = 0.1

```

# Load data
Load patient information from the Nature Cancer paper repo (https://github.com/FertigLab/J15221_trial_paper).

```{r echo=FALSE, message = FALSE, warning = FALSE}
patInfoLink = "https://raw.githubusercontent.com/FertigLab/J15221_trial_paper/refs/heads/main/input_tables/table_24patient_info.csv"

patInfo = read.csv(patInfoLink, row.names = 1, stringsAsFactors = F)

# read sample into and merge with patient info
sampAnnot = read.csv(paste0(inputFolder,"diversity_neoantigen_table.csv"), row.names = 1)

sampAnnot = cbind(sampAnnot, patInfo[sampAnnot$patientID, c("Best_Resp", "biopsy_site")])

# add timepoint
sampAnnot = sampAnnot %>% mutate(Timepoint = Time)

```

# CIBERSORTx proportions

The analysis was performed using the CIBERSORTx web portal (https://cibersortx.stanford.edu/) with the LM22 signature matrix. The results were downloaded and processed in R.


```{r echo=FALSE, fig.width = 8, message = FALSE, warning = FALSE}
#=============================== 
# read in CIBERSORT proportions
ciberRes = read.csv(file = paste0(inputFolder,'TME/CIBERSORTx_proportions.csv'), row.names = 1)

# subset to samples that we agreed to use for further analysis
ciberRes = ciberRes[rownames(sampAnnot),1:22]

```

## Heatmap

```{r echo=FALSE,  fig.width = 10, message = FALSE, warning = FALSE}
#===============
# make heatmaps

# find cell types that don't show a lot of variation
ciberSD1 = apply(ciberRes,2,sd)
cellTypes = names(ciberSD1)[which(ciberSD1 > 0.01)]

  #=====================
  # plot only cell type with some variation
  dat = t(ciberRes[,cellTypes])
  annot = data.frame(Response = factor(sampAnnot[,"Best_Resp"], levels = c("CR","PR","SD","PD")),
                     RECIST=sampAnnot[,'RECIST'],
                    Biopsy_site = sampAnnot[,'biopsy_site'],
                       Type = sampAnnot[,'Subtype'],
                          Timepoint=sampAnnot[,'Timepoint']
                       )
  rownames(annot) = rownames(sampAnnot)

  annot_color = list(Timepoint = timeCol, Type = subtypeCol, RECIST = recistCol, Biopsy_site = biopsyCol, Response = respCol)
  
# all samples
#  pheatmap(dat,annotation_col = annot, main = 'CIBERSORT. All samples', scale = "none", annotation_colors = annot_color, labels_col = sampAnnot[colnames(dat),"pubID"])
  
# ordered by patient and time
  ord = rownames(sampAnnot)[order(sampAnnot$RECIST,sampAnnot$patientID,sampAnnot$Timepoint)]
  pheatmap(dat[,ord],annotation_col = annot[ord,], main = 'CIBERSORT. All samples', 
           scale = "none", annotation_colors = annot_color, 
           labels_col = sampAnnot[ord,"patientID"], cluster_cols = FALSE, 
           gaps_col = c(3,5,7,8,10,11,13,14,16,18,21,22,24,26,28,30,32,35))

```

## p-values

```{r echo=FALSE, fig.width= 10, warning=FALSE, message=FALSE}

# cibersort proportions
df = data.frame(ciberRes, 
                sampAnnot[rownames(ciberRes), ],
                check.names = F)
  # calculate p-values
#  res_ciber = t(sapply(colnames(ciberRes), getWilcox, df))
  res_ciber = t(sapply(colnames(ciberRes), getWilcoxDelta, df))
  
  res_ciber = round(res_ciber,3)
# table with p-values
  datatable(res_ciber)
  
  # save the tables with p-values for the paper supplement
  write.csv(res_ciber, file = paste0(outputFolder,'cibersort_pvalues.csv'))


```

## Boxplots responders vs non-responders

```{r echo=FALSE, fig.width=10}
# get significant by time point
sig = rownames(res_ciber)[which(res_ciber[,"pValue_B_RvsNR"] < 0.05)]

##RECIST (Responders vs. Nonresponders)
for( i in sig)
{
df = data.frame(cell_proportion = as.numeric(ciberRes[,i]), 
                sampAnnot[rownames(ciberRes), ],
               check.names = F)

p= ggplot(df, aes(x=RECIST, y=cell_proportion)) + 
  geom_boxplot() + 
  ylab(i) + 
  stat_compare_means() + 
  geom_point() + 
#  scale_y_log10()+
#  scale_color_manual(values = subtypeCol) +
  facet_wrap("Timepoint")+ 
  theme_bw()

print(p)
}

```

## Boxplots for Monocytes

Plot treatment effect between responders and non-responder (after entinostat)

```{r echo=FALSE, fig.width=4, fig.height=4, warning=FALSE, message=FALSE}
df = data.frame(Monocytes = as.numeric(ciberRes[,"Monocytes"]), 
                sampAnnot[rownames(ciberRes), ],
                check.names = F)

# per patient matrix with time point columns
mat <- data.frame(matrix(nrow = length(unique(df$patientID)), ncol = 3, 
                         dimnames = list(unique(df$patientID),
                                         levels(factor(df$Timepoint)))),
                  check.names = F)
# iterate time and fill in matrix
for(i in levels(factor(df$Timepoint)))
  {
    d1 = df %>% filter(Timepoint == i)
    mat[d1$patientID, i] = d1[,"Monocytes"]
  }
# add response
m = df %>% select(patientID,  RECIST) %>% filter(!(duplicated(patientID)))
rownames(m) = m$patientID
mat$RECIST = m[rownames(mat),"RECIST"]
# create treatment effect (calculate delta)
# substract baseline from post run-in and week 8 
mat$delta_T1_B =  mat[,2]-mat[,1]

# delta for responders vs non-responders 
ggplot(mat, aes(x=RECIST, y=delta_T1_B)) + 
  geom_boxplot() + 
  ylab("delta(C1D1 - Baseline)") + 
  ggtitle("Entinostat treatment effect") +
  stat_compare_means() + 
  geom_point(size = 2) + 
  theme_bw()

```

# TME gene sets
## p-values

```{r echo=FALSE, fig.width = 8, message = FALSE, warning = FALSE}
# read in scores
scores = read.csv(paste0(inputFolder,'TME/TME_gene_sets_scores.csv'), row.names = 1, check.names = F)

df = data.frame(t(scores), 
              sampAnnot[colnames(scores), c("RECIST","Subtype","patientID","Time")],
              check.names = F)
  # calculate p-values
  res = t(sapply(rownames(scores), getWilcoxDelta, df))
  
  res = round(res,3)
# table with p-values
  datatable(res)
  
write.csv(res, file = paste0(outputFolder,"TME_gene_sets_p-values.csv"), quote = F)

```

## Significant sets in responders vs non-responder

### at C1D1

```{r echo=FALSE, fig.width = 8, message = FALSE, warning = FALSE}
sig_t1 = rownames(res)[which(res[,"pValue_T1_RvsNR"] < pThresh)]

# order by patient and time
ord_all = rownames(sampAnnot)[order(sampAnnot$RECIST,sampAnnot$Best_Resp, sampAnnot$Subtype, sampAnnot$patientID,sampAnnot$Timepoint)]
# C1D1 only
ord = grep("C1D1",ord_all, value = T)
# heatmap
plotHA(sampAnnot[ord,],scores[sig_t1,ord], scale = F, title = "TME gene sets. C1D1", col = heatmapCol, column_order = ord, column_labels  = gsub("_C1D1","",ord))

```

### at Week 8

```{r echo=FALSE, fig.width=8, eval=TRUE, warning=FALSE, message=FALSE}
# get significant by time point
sig_t2 = rownames(res)[which(res[,"pValue_T2_RvsNR"] < pThresh)]


# Week 8 only
ord = grep("Week_8",ord_all, value = T)
plotHA(sampAnnot[ord,],scores[sig_t2,ord], scale = F, title = "TME gene sets. Week 8", col = heatmapCol, column_order = ord,
          column_labels  = gsub("_Week_8","",ord) )


```

# R libraries and versions

```{r echo=FALSE, message = FALSE, warning = FALSE}

sessionInfo()

```
